### 2015-03-31 paper/chemokine_main_revised.tex
RSI Submission

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we focus on T cell navigating from the lymph nodes, through the vascular system to the foci, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.

We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models. As plaques expand in size, infected regions become more isolated, making it difficult for T cells to cover the infection efficiently.  Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.  However, the model parameters that collectively determine how efficiently T cells find and clear infected cells are also significant in the model.  These model observations provide explanation of the pandemic H1N1 dynamics that have not been well explained by previous studies.

@@

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood..Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging. We address this challenege with computational modeling, which allows us to study the relative contributions of different aspects of the host immune response, and to account for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Here, we present a comprehensive agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models of to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008}.  We use a combination of in vitro generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection to parameterize our model.  Once parameterized with these data, the model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.  These three strains provide a range of documented pathogenicities abd \textit{in vitro} replication efficiencies [TODO citation(s) with pathogenicity data of 3 strains] \cite{Mitchell2011}, which we study in the model.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

[TODO rewrite to be more specific] The model design allows us to compare between influenza strains how the combination of viral and immune system characteristics affect the strength of the given infection.  The results give us specific insights into how these features [TODO list some] determine the pathogenicity of the different viruses.

The spatial nature of the model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the host cells' chemokine response may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis isolated the effects of each parameter in the model.  Based on this analysis, each parameter was classified as critical or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on model results.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus, confirming that the higher replication rate of 2009 pH1N1 influenza overwhelmed contributions of the chemotactic signals, leading to uncontrolled \textit{in silico} infection.

% fred used 'obscurred' instead of overwhelmed

@@

Our modeling methodology follows three steps.  First, we define the system we are going to model.  Second, we build the model with this definition in mind.  Finally, we parameterize the model using empirical data, values from literature, and computational techniques.

@@

The model is implemented using the CyCells software \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).  Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models [TODO add Catherine ref?] \cite{Miller2003, Allan2006, Ingulli2009}.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).

@@

[TODO cite Catherine or ourselves] In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.

@@

Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}. These values were taken from the literature when available.  Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.  All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.  [TODO number of model parameters?]

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).  T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2). T cell production is assumed to be constant after day 5 [TODO reference or justification for constant rate].

@@

We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).  
The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs. Consistent with earlier results \cite{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).

In all studied strains, virus expansion declines at four days post infection (Fig.~\ref{fig:variance}), due to IgM activation.  Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five post infection [TODO p.i. for post infection].  aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six post infection, pH1N1 replication dominates the IgM and T cell response.  These results agree with [TODO "previously reported" + citation] \textit{in vivo} differences among the three different strains.

@@

% Spatial effects of viral replication and chemokine diffusion play an important role in both the spread and clearance of infections.  [TODO maybe place in intro]

@@

Plaque growth in the model is illustrated in Figure \ref{fig:cycells}.  Because the virus particles are an order of magnitude larger than the chemokine moledules, they diffuse more slowly, according to the Stokes Einstein equation.  However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).  In the model these countervailing forces produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}). Before day 4 the plaque is dominated by active (incubating and secreting) cells, with a small percentage of dead cells (Panel A). Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B). By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells (Panel C).  Alhtough not visible in the figure, plaque size and T cell number changes coninuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).

% [TODO label fig. 5 with A B C...]

@@

Finally, the spatial nature of our model reveals that regions of concentrated chemokine diffuse more slowly than the ever-expanding infected cell and virus locations, thus misdirecting the chemotaxing T cells.  It takes time for infected cells to begin producing chemokine, while the preexisting pockets of high chemokine density are slow to decay.  Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.  This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse..

@@

We conducted a sensitivity analysis of sixteen model parameters to identify those that are most important for determining model outcomes.  We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.  T cell radius is known \cite{abbas2011cellular} and does not factor strongly into the spatial model's implementation.  T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine ([TODO reference supplement]).  The two onset parameters are known \cite{Diamond2003}.  Viral production is an independent variable, as discussed.above, and determined by each of the three strains \cite{Mitchell2011}.

% The results were examined individually and are discussed in depth in the supplement.

Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.  Results were compared for the size of the peak infection and the number of ucleared infected cells at day 10 p.i.. Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{mostly stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold, \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.

Table \ref{tab:sensitivity} shows that five parameters are sensitive: IgM strength, infection sensitivity, viral decay rate, viral diffusion, and the T cell secretion rate.  In addition, section [TOOD reference] shows that viral production rate is a sensitive parameter.  Four of the parameters (infection sensitivity, viral decay rate, viral diffusion rate, and viral production rate) are directly related to viral efficacy in isolation.  Further, IgM strength, as coded in the model, directly affects viral decay rate and can be included in this group.  The T cell secretion rate is the only sensitive parameter that does not directly control virus kinetics.  None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model.  This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.

\subsection*{Window of Vulnerability}

An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s (Table~\ref{tab:strains}).  Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes. viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles, which can infect new cells.  Thus, T cells alone cannot contain the pH1N1 infection in our model.  By contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the Avian strain this number is 0.2.  % Thus, viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.

Even if apoptosis was instantaneous (apoptosis time equals 0) the cell would still secrete virus for 10 minutes, producing 3 new virions (pH1N1).  We tested this effect in the model by setting both relevant parameters to zero, and indeed with this assumption all three strains were cleared by the T cell response (Fig.~\ref{fig:instantkill}).  We now ask how much T cell delay the model can tolerate and still clear a given viral strain.  To answer this, we formulate the basic equations of the model in terms of $R=0$, the viral replication rate, and solve for the case where $R_0 < 1$.

\begin{equation}
\lim_{t \to \infty} R_0 = \frac{p}{2} * (T_{arr} + T_{kill} + T_{apop})
\end{equation}

% $R_0$ can simplistically be defined as the viral production rate multiplied by the time of production.  Using this definition, the claculated $R_0$ of the sH1N1 virus is 1.6, suggesting growth, yet the actual infection is eventually cleared  by day 10 in our model.  This can be attributed to the spatial effects of the ever-expanding ring of infection.

%[TODO justify using the limit of p/2]

This equation accounts for spatial effects of viral diffusion in the following way.  Some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.  This effect is minimal in the beginning when the edge of the plaque is tightly curved.  As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus diffuses over healthy cells and 50\% over infected or dead cells.  We correct for this in the equation by dividing $p$ by 2.  [TODO This correction ignores the viral gradient in the plaque which pushes the new virus particles slightly outwards.]  Equation 1 explains how sH1N1 can be cleared in the model even though it produces 1.6 viral particles during the vulnerability window.  Equation 1 calculates the effective $R_0$ for sH1N1 to be 0.8.

------------ STOP HERE, FIX BELOW  ----------------

Where $T_{arr}$ is the time it takes for a T cell to arrive at the location of an infected cell, $T_{kill}$ is the time it takes for the T cell to induce apoptosis, and $T_{apop}$ is the apoptosis time.  Note that this equation calculates the number of virions that may be produced by an infected cell, but omits the number of cells that can subsequently be infected by the virus.  A quick calculation of a virion's expected life span in the presence of IgM (2.4 hours) combined with the expected time it takes to infect a healthy cell (2 hours), results in a value greater than one.  Given that virus is removed from the model when a cell becomes infected, each virus is limited to infecting one target cell, and thus this quantity does not affect the equation for $R_0$.  Given these assumptions, we can then solve for $T_{arr}$ where $R_0 < 1$:

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Viral clearance depends in part on the ability of T cells to locate the infection in time to prevent uncontrolled spread of the virus.  We use an agent-based spatial model to quantify how efficiently T cells locate and control different strains of influenza in the lung.  We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature.  The spatial model reveals challenges for T cell recruitment not apparent in standard differential equation models.  The model shows that virus kinetics largely determine the course of the infection, while T cell chemokine parameters have less effect.

%Further analysis of the model identifies the viral parameters as most significant and suggests possible mitigation strategies.

% ---- 300 word abstract

% Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we focus on T cell navigating from the lymph nodes through the vascular system to the infection, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.

%We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models. As plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.  Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.  Finally, we use the model parameters to determine an upper bound on T cell search time in the model.

% ---- end 300 word abstract

@@

\section*{Keywords}

Agent-Based Model, Systems Biology, Immunology, Computational Biology, Virology

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.  Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging. We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.  We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.  The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.  These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

% The model design allows us to compare between influenza strains how the T cell response is affected by the viral and immune system characteristics.  The results show that efficient T cell search is crucial to viral clearance in the model and that the search process is dependent on more than just the replication rate of the virus, including the spatial landscape of the chemokine gradient and the location and density of virus-secreting cells.

The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.  Based on this analysis, each parameter was classified as sensitive or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.

% fred used 'obscured' instead of overwhelmed

@@

% Our modeling methodology follows three steps.  First, we define the system we are going to model.  Second, we build the model with this definition in mind.  Finally, we parameterize the model using empirical data, values from literature, and computational techniques.

@@

The model is implemented using the CyCells software \cite{Warrender2006} (\texttt{https://github.com/drewlevin/cycells}), a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).  Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \cite{Miller2003, Allan2006, Ingulli2009}.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \cite{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.

@@

Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}. These values were taken from the literature when available.  Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.  All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).  T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2). T cell production is assumed to be constant after day 5 \cite{MartIn-Fontecha2003}.

@@

% We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).  The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs. Consistent with earlier results \cite{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).

We estimated variance across runs by calculating the standard deviation at each time point.  For aH5N1, the maximum standard deviation was 37, which occurred at day 5.05 where the mean value was 259.  For sH1N1, the maximum standard deviation was 242, which occurred at day 4.93 where the mean value was 2,607.  For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018.

In all studied strains, virus expansion declines at four days p.i. (Fig.~\ref{fig:variance}), due to IgM appearance.  Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five p.i.  aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six p.i., pH1N1 replication dominates the IgM and T cell response.  These results agree with previously reported \textit{in vivo} differences among the three different strains \cite{Mitchell2011}.

% Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.

@@

@@

Plaque growth in the model is illustrated in Figure \ref{fig:cycells}.  Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly, according to the Stokes-Einstein equation.  However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).  In the model these countervailing forces produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}). Before day 4 the plaque is dominated by active (incubating and secreting) cells, with a small percentage of dead cells (Panel A). Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B). By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells (Panel C).  Although not visible in the figure, plaque size and T cell number changes continuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).

@@

Finally, the spatial nature of our model reveals that regions of concentrated chemokine diffuse more slowly than the ever-expanding infected cell and virus locations, thus misdirecting the chemotaxing T cells.  It takes time for infected cells to begin producing chemokine, while the preexisting pockets of high chemokine density are slow to decay.  Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.  This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.

@@

We conducted a sensitivity analysis of sixteen model parameters to identify those that are most important for determining model outcomes.  We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.  T cell radius is known \cite{abbas2011cellular} and does not factor strongly into the spatial model's implementation.  T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine (S2.3).  The two onset parameters are known \cite{Diamond2003}.  Viral production is an independent variable, as discussed above, and determined by each of the three strains \cite{Mitchell2011}.

Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.  Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.  Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.

Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.  In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.  The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.  Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.  None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.  This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.

\subsection*{Window of Control}

Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).  We asked whether T cells could play a role by assessing multiple factors that control the overall T cell response.  Here, we define the window of control as the maximum time virus-secreting cells may produce virus while still allowing the infection to be cleared.  This window can be defined as the combination of the the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).

An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s (Table~\ref{tab:strains}).  Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes of T cell arrival, viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles that can infect new cells.  In contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the aH5N1 strain this number is 0.2.  % Thus, viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.

Even if T cell arrival and apoptosis time were instantaneous, a cell would still secrete virus for 10 minutes ($T_{kill}$), producing 3 new virions in the case of pH1N1.  Only in the case where both $T_{kill}$ and $T_{apop}$ are simultaneously reduced can the pandemic infection be cleared.  We confirmed this effect in the model by first setting both relevant parameters (apoptosis time and T cell kill time) to zero, and as expected all three strains were cleared (Fig.~\ref{fig:instantkill}).  We now ask how much T cell delay the model can tolerate and still clear a given viral strain.  To answer this, we formulate the basic equations of the model in terms of $R_0$, the viral replication rate, and solve for the case where $R_0 < 1$.

\begin{equation}
R_0 = p \times (T_{arr} + T_{kill} + T_{apop}) \times E
\end{equation}

\noindent where $p$ is the secretion rate of the virus and $E$ is viral effectiveness, i.e. the proportion of virus that infects cells.  Equation 1 assumes each free virus particle can infect no more than one healthy epithelial cell.

%%%
% Note that this equation calculates the number of virions that may be produced by an infected cell, but omits the number of cells that can subsequently be infected by the virus.  A quick calculation of a virion's expected life span in the presence of IgM (2.4 hours) combined with the expected time it takes to infect a healthy cell (2 hours), results in a value greater than one.  Given that virus is removed from the model when a cell becomes infected, each virus is limited to infecting one target cell, and thus this quantity does not affect the equation for $R_0$.

% $R_0$ can simplistically be defined as the viral production rate multiplied by the time of production.  Using this definition, the claculated $R_0$ of the sH1N1 virus is 1.6, suggesting growth, yet the actual infection is eventually cleared  by day 10 in our model.  This can be attributed to the spatial effects of the ever-expanding ring of infection.  
%%%
